---
title: "Setting Up DESCRIPTION and dependencies.R"
subtitle: "Instructions for Mass Dynamics Custom R Workflows"
author: "Mass Dynamics"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

# Overview

This tutorial explains how to create and maintain `DESCRIPTION` and `dependencies.R` for your Mass Dynamics custom R workflow. The approach depends on whether you are building an **R package** or using a standalone **R script**.

------------------------------------------------------------------------

# Case 1: No DESCRIPTION — Creating an R Package from Scratch

If you do not yet have an R package, start by creating one.

## Step 1: Create the package structure

**Option A — Using `usethis` (recommended):**

```{r}
# Install usethis if needed
install.packages("usethis")

# Create a new package (run from the directory where you want the package folder)
usethis::create_package("MyWorkflow")
```

This creates a folder with `DESCRIPTION`, `NAMESPACE`, `R/`, and basic structure.

**Option B — Using RStudio:**

Use **File → New Project → New Directory → R Package**, then enter the package name.

**Option C — Manual:**

Create a folder and add a minimal `DESCRIPTION` file:

```         
Package: MyWorkflow
Title: My Custom Workflow
Version: 0.1.0
Author: Your Name
Maintainer: Your Name <you@example.com>
Description: Custom R workflow for Mass Dynamics.
License: MIT + file LICENSE
Imports:
    R (>= 4.0)
```

## Step 2: Add your R code

Place your workflow functions in `R/` (e.g. `R/my_workflow.R`). Use `@importFrom` in roxygen2 comments to declare dependencies:

```{r}
#' Run my workflow
#'
#' @param data Input dataframe
#' @importFrom dplyr filter
#' @importFrom limma normalizeBetweenArrays
#' @export
my_workflow <- function(data) {
  # ...
}
```

## Step 3: Update DESCRIPTION from your code

Use `attachment` to keep `DESCRIPTION` in sync with what you use:

```{r}
install.packages("attachment")
attachment::att_amend_desc()
```

This scans `R/`, `tests/`, and vignettes and adds any used packages to `Imports` or `Suggests`.

## Step 4: Create dependencies.R

`dependencies.R` installs packages *before* your package is built during the Docker image build.

**CRAN packages:**

```{r}
install.packages("dplyr", repos = "https://cloud.r-project.org")
install.packages("tidyr", repos = "https://cloud.r-project.org")
```

**Bioconductor packages (e.g. limma):**

```{r}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager", repos = "https://cloud.r-project.org")

BiocManager::install("limma")
```

**Build tools (if needed for install.R):**

```{r}
install.packages("devtools", repos = "https://cloud.r-project.org")
```

Place `dependencies.R` in the **root** of your repository (same level as `DESCRIPTION`).

------------------------------------------------------------------------

# Case 2: DESCRIPTION Already Exists — You Have an R Package

If you already have an R package with a `DESCRIPTION` file, use it as the source of truth and keep it in sync with your code.

## Step 1: Sync DESCRIPTION from your code

Run this regularly (e.g. after adding new `library()` calls or `@importFrom`):

```{r}
library(attachment)
att_amend_desc()
```

This updates `Imports` and `Suggests` from:

-   `R/` scripts
-   `tests/` scripts
-   `@importFrom` / `@import` in roxygen2
-   `pkg::fun()` calls

Use `extra.suggests = c("pkgdown", "testthat")` if you use dev tools. Save the config for next time:

```{r}
att_amend_desc(extra.suggests = c("pkgdown", "testthat"), update.config = TRUE)
```

## Step 2: Create or update dependencies.R

**When DESCRIPTION exists**, `dependencies.R` should install:

1.  **Build tools** — `devtools` (for `install.R`), optionally `BiocManager`
2.  **Bioconductor packages** — anything from Bioconductor (e.g. `limma`, `DESeq2`) via `BiocManager::install()`
3.  **Other pre-install needs** — packages that must be present before `devtools::install()` runs

CRAN packages in `DESCRIPTION` are typically installed when `devtools::install()` runs, but Bioconductor packages often need to be installed first.

**Example dependencies.R for an R package:**

```{r}
# Build tools
install.packages("devtools", repos = "https://cloud.r-project.org")

# Bioconductor packages (check DESCRIPTION Imports for these)
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager", repos = "https://cloud.r-project.org")

BiocManager::install("limma")   # Add any Bioconductor packages from your DESCRIPTION
```

**Optional: generate a draft from DESCRIPTION**

```{r}
library(attachment)
create_dependencies_file(path = "DESCRIPTION", to = "dependencies.R", open_file = FALSE)
# Review and edit the output — Bioconductor packages may need manual BiocManager::install() calls
```

------------------------------------------------------------------------

# Case 3: R Script Only — No Package, No DESCRIPTION

If you use a standalone R script (no package structure), you do not need `DESCRIPTION` or `install.R`.

## Create dependencies.R

`dependencies.R` is your only dependency list. Install every package your script uses:

```{r}
# CRAN packages
install.packages(c("dplyr", "tidyr", "readr"), repos = "https://cloud.r-project.org")

# Bioconductor packages
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager", repos = "https://cloud.r-project.org")

BiocManager::install("limma")
```

Place `dependencies.R` in the root of your repository. The Dockerfile will run it before copying and running your R script.

------------------------------------------------------------------------

# Quick Reference

| Scenario               | DESCRIPTION                                         | dependencies.R                          | install.R             |
|------------------------|-----------------------------------------------------|-----------------------------------------|-----------------------|
| **New R package**      | Create with `usethis::create_package()` or manually | Create with CRAN + BiocManager installs | `devtools::install()` |
| **Existing R package** | Keep in sync with `att_amend_desc()`                | Install build tools + Bioconductor      | `devtools::install()` |
| **R script only**      | Not needed                                          | Install all packages used               | Not needed            |

------------------------------------------------------------------------

# Tips

-   **Version constraints** — Add them in `DESCRIPTION` (e.g. `limma (>= 3.42.2)`). `dependencies.R` usually installs latest unless you pin versions explicitly.
-   **System dependencies** — If the Docker build fails with missing libraries (e.g. libxml2), add `dependencies.sh` to install them. See the [README](../../README.md) for details.
-   **Tracking over time** — For large packages, run `att_amend_desc()` before releases to ensure `DESCRIPTION` matches the codebase.
